---
title: "Species Richness"
authors: "Occidental Computational Biology Lab"
date: "Updated: 2025-09-16"
output: pdf_document
---

This file runs the species richness models

```{r}

# TODO: add comments

library(dplyr)
library(tidyr)
library(ggplot2)
library(forcats)
library(iNEXT)
library(googlesheets4)

```


## Load Photo Database Files

```{r}

load_file <- function(file, city, camera_sites_file) {
  photos <- file %>%
    rename("Species" = "commonName")
  
  if (city == "PACA") {
    camera_sites_file <- camera_sites_file %>%
      mutate(
        locAbbr = sub("^[^-]+-", "", locAbbr),
        locAbbr2 = sub("^[^-]+-", "", locAbbr2)
      )
  }
  
  photos <- photos %>%
    left_join(
      camera_sites_file,
      by = c("City", "locAbbr")
    ) %>%
    drop_na(locAbbr2)
  
  return(photos)
}

```


```{r}

filter_threshold <- function(file, threshold) {
  file <- file %>%
    filter(maxDetectionConf >= threshold)
  return(file)
}

```

## Convert Timestamp Column to Datetime Format

```{r}

convert_timestamp <- function(file) {
  file$TimestampUTC <- file$Timestamp
  file$TimestampUTC <- as.POSIXct(file$TimestampUTC,
                                           format="%Y-%m-%dT%H:%M:%OSZ", tz="UTC")
  
  # Extract month
  file$Month <- format(file$TimestampUTC, format="%m")
  
  # Extract year-month
  file$YearMonth <- format(file$TimestampUTC, format="%Y-%m")
  
  return(file)
}

```

## Split photos into species detections
- Some photos have multiple species detections. This splits them into separate rows, one for each species detected in the photo.

```{r}

split_photos <- function(file) {
  # Step 1: Separate out rows that contain underscores
  with_underscores <- file %>%
    filter(grepl("_", Species))
  
  # Step 2: For each of those rows, split Species into multiple rows
  split_rows <- with_underscores %>%
    mutate(row_id = row_number()) %>%  # add ID to track original rows
    separate_rows(Species, sep = "_")  # split into multiple rows by "_"
  
  # Step 3: Keep rows without underscores
  without_underscores <- file %>%
    filter(!grepl("_", Species))
  
  # Step 4: Combine all together
  spp_detections <- bind_rows(without_underscores, split_rows) %>%
    select(-row_id)  # remove temp ID column if desired
  
  # TODO: is this necessary?
  rm(without_underscores)
  rm(split_rows)
  
  return(spp_detections)
}

```

## Bind Species ID Data to Species Detection Data
- This allows us to categorize species into taxonomic groups

```{r}

combine_species_data <- function(file, species_groups_file) {
  # TODO: update species file format?
  # Change format such that Subspecies (commonName in df) maps to SpeciesGroup
  species_map <- species_groups_file %>%
    pivot_longer(
      cols = -1,
      names_to = "SubCol",
      values_to = "Subspecies"
    ) %>%
    filter(!is.na(Subspecies), Subspecies != "") %>%
    rename(SpeciesGroup = 1) %>%
    select(SpeciesGroup, Subspecies)
  
  # Map commonName to SpeciesGroup value
  file <- file %>%
    left_join(species_map, by = c("Species" = "Subspecies")) %>%
    filter(SpeciesGroup != "DELETE" | is.na(SpeciesGroup)) %>%
    mutate(Species = coalesce(SpeciesGroup, Species)) %>%
    filter(!is.na(Species)) %>%
    select(-SpeciesGroup)
  
  print(unique(file$Species))
  
  return(file)
}

```

## Filter out Human and Empty Photos
```{r}

filter_species <- function(file) {
  file <- subset(file,
                   Species != "Empty" &
                   Species != "Human" &
                   Species != "Unknown" &
                   Species != "Bird" &
                   Species != "Domestic cat" &
                   Species != "Domestic dog" &
                   Species != "Horse" &
                   Species != "Insect" &
                   Species != "Lizard" &
                   Species != "Rodent" &
                   Species != "Snake" &
                   Species != "Amphibians" &
                   Species != "Turtle"
                   # TODO: add more species to exclude
                 )
  
  return(file)
}

```

## Species Stats
- Species Richness Raw: Total number of different species
- Note this includes cannot ID categories

```{r}

total_species <- function(file) {
  print(length(unique(file$Species)))
}

```

## Create Directories for Each Camera Site

```{r}

create_dirs <- function(file, output_path) {
  site_list <- c("ALL", unique(file$locAbbr2))
  for(i in site_list){
    dir.create(paste0(output_path, "Sites/", i), recursive = TRUE, showWarnings = FALSE)
  }
  
  return(site_list)
}

```

Bar chart of number of observations of each species

```{r}

num_observations <- function(site_list, file, output_path) {
  for(i in site_list){
    plot_data <- if(i == "ALL") file else subset(file, locAbbr2 == i)
    
    ggplot()+
      geom_bar(data=plot_data, aes(x=fct_infreq(Species)))+
      theme_bw(base_size=6)+
      coord_flip()+
      xlab("Species")+
      ylab("Number of Photos")
    
    ggsave(paste0(output_path, "Sites/", i, "/species_freq_all.jpg"), units = "in", dpi = 600)
  }
}

```

###########################

## Species Accumulation Curve

<https://rpubs.com/brouwern/iNEXTvVEGAN> <https://www.researchgate.net/post/How_do_I_generate_a_species_accumulation_curve_based_on_sampling_effort_for_a_camera-trap_study_using_R>

## By abundance

```{r}

species_accumulation_curve <- function(site_list, file, output_path) {
  for(i in site_list){
    print(i)
    species.vec1 <- if(i == "ALL") file else subset(file, locAbbr2 == i)
    
    if(nrow(species.vec1) > 0) {
      species.vec <- table(species.vec1$Species) ##abundance
      
      # TODO: look into Season as variable
      abundance.out.inext <- iNEXT(as.vector(species.vec), q=0, datatype="abundance")
      
      results <- abundance.out.inext$AsyEst
      results$`Total Observations` <- sum(species.vec)
      
      write.csv(results, paste0(output_path, "Sites/", i, "/species_richness_abun.csv"), row.names=T, quote=F)
      
      ggiNEXT(abundance.out.inext, type=1, facet.var="None")+
        theme_bw()+
        scale_x_continuous(expand = c(0, 0)) + #, limits = c(0, 62)
        ylim(0, 20) + #scale_y_continuous(expand = c(0, 20)) + #, 
        xlab("Number of spp_detections") +
        #ylab("Species diversity") +
        #labs(title= "Species Accumulation Curve")+
        scale_color_manual(values="mediumorchid")+
        scale_fill_manual(values="mediumorchid")+ 
        theme(legend.title=element_blank(), 
              legend.position = "bottom",
              strip.background = element_blank(),
              axis.title.y = element_text(size = 12),
              axis.title.x = element_text(size = 12))
      
      ggsave(paste0(output_path, "Sites/", i, "/species_richness_abun.jpg"), height=5, width=8, units = "in", dpi = 600)
    }
  }
}

```


```{r}

combine_results <- function(output_path) {
  # Recursively find all files named "species_richness_abun.csv"
  files <- list.files(path = output_path, 
                      pattern = "^species_richness_abun\\.csv$", 
                      recursive = TRUE, 
                      full.names = TRUE)
  
  # Read all files, add a Site column, and combine them
  all_data <- do.call(rbind, lapply(files, function(f) {
    # Extract site name from parent folder
    site <- basename(dirname(f))
    
    # Read CSV file
    df <- read.csv(f, stringsAsFactors = FALSE)
    
    # Add Site column
    df$Site <- site
    
    return(df)
  }))
  
  return(all_data)
}

```

## Merge Imperviousness with Richness

```{r}

merge_impervious <- function(file, richness_file, output_path) {
  camera_sites <- file %>%
    select(locAbbr2, Impervious) %>%
    distinct(locAbbr2, .keep_all = TRUE)

  site_data <- merge(richness_file, camera_sites, by.x = "Site", by.y = "locAbbr2")
  write.csv(site_data, paste0(output_path, "richness.csv"), quote=F, row.names=F)
  
  return(site_data)
}

```


```{r}

evaluate_results <- function(file, output_path) {
  ggplot(data=file, aes(x=Impervious, y=Estimator))+
    geom_point()+
    stat_smooth(method="lm")+
    facet_wrap(.~X, scales="free_y")+
    theme_bw()
  
  ggsave(paste0(output_path, "Impervious x Diversity.jpg"))
  
  ggplot(data=subset(file, X == "Species Richness"), aes(x=Impervious, y=Estimator))+
    geom_point()+
    stat_smooth(method="lm")+
    facet_wrap(.~X, scales="free_y")+
    theme_bw()
  
  ggsave(paste0(output_path, "Impervious x Richness.jpg"))
  
  lm.out <- lm(Estimator ~ Impervious, data=file, subset = X == "Species Richness")
  
  coef_table <- summary(lm.out)$coefficients
  coef_df <- as.data.frame(coef_table)
  coef_df$Term <- rownames(coef_df)
  rownames(coef_df) <- NULL
  colnames(coef_df) <- c("Estimate", "Std_Error", "t_value", "p_value", "Term")
  coef_df <- coef_df %>%
    select(Term, Estimate, Std_Error, t_value, p_value)
  
  write.csv(coef_df, paste0(output_path, "lm_results.csv"), quote=F, row.names=F)
}

```

## Iterate over all cities and thresholds, using functions defined above

```{r}

cities <- c(
  "AUTX",
  "CHIL",
  "ICIA",
  "ININ",
  "JAMS",
  "PACA",
  "BOMA",
  "SLMO",
  "TAWA",
  "URIL"
)
thresholds <- c(0, 0.1, 0.2, 0.3, 0.4, 0.5, 0.6, 0.7, 0.8, 0.9, 0.99)

camera_sites <- read.csv("../Input Data/camera_sites_with_impervious.csv")

# Get species group mapping from Google Sheets "Species Groups"
species_groups <- range_read("https://docs.google.com/spreadsheets/d/1czht5zz8dRII25loY9RI9efTi6vIsvsioSqXBgIwuvg/edit?gid=0#gid=0", trim_ws = FALSE)

richness_path <- "../Richness/"

all_photos <- read.csv(paste0("../Input Data/multi_city_data_cleaned.csv"))

for (city_name in cities) {
  city_photos <- all_photos %>%
    filter(City == city_name)
  for (threshold_value in thresholds) {
    output_path_name <- paste0(richness_path, city_name, "/Conf. ", threshold_value, "/")
    
    print(paste("City:", city_name))
    print(paste("Threshold:", threshold_value))

    photos <- load_file(city_photos, city_name, camera_sites)
    filtered <- filter_threshold(photos, threshold_value)
    converted <- convert_timestamp(filtered)
    split <- split_photos(converted)
    combined <- combine_species_data(split, species_groups)
    exclude <- filter_species(combined)
    total_species(exclude)
    sites <- create_dirs(exclude, output_path_name)
    num_observations(sites, exclude, output_path_name)
    species_accumulation_curve(sites, exclude, output_path_name)
    richness <- combine_results(output_path_name)
    with_impervious <- merge_impervious(exclude, richness, output_path_name)
    evaluate_results(with_impervious, output_path_name)
  }
}

```

# Collect all results within city across thresholds

```{r}

for (city_name in cities) {
  city_results <- data.frame(
    City = character(),
    Threshold = numeric(),
    Term = character(),
    Estimate = numeric(),
    Std_Error = numeric(),
    t_value = numeric(),
    p_value = numeric()
  )
  
  city_richness_results <- data.frame(
    City = character(),
    Threshold = numeric(),
    Site = character(),
    X = numeric(),
    Estimator = numeric(),
    Est_s.e. = numeric(),
    X95..Lower = numeric(),
    X95..Upper = numeric(),
    Total.Observations = numeric(),
    Impervious = numeric()
  )
  
  for (threshold_value in thresholds) {
    output_path_name <- paste0(richness_path, city_name, "/Conf. ", threshold_value, "/")
    
    # Read the linear model results for this threshold
    results_path <- paste0(output_path_name, "lm_results.csv")
    
    richness_file_path <- paste0(output_path_name, "richness.csv")
    
    if (file.exists(results_path) == F) {
      next
    }
    
    if (file.exists(richness_file_path) == F) {
      next
    }
    
    cur_file <- read.csv(results_path)
    
    cur_file$City <- city_name
    cur_file$Threshold <- threshold_value
    
    city_results <- rbind(city_results, cur_file)
    
    cur_richness_file <- read.csv(richness_file_path)
    
    cur_richness_file$City <- city_name
    cur_richness_file$Threshold <- threshold_value
    
    city_richness_results <- rbind(city_richness_results, cur_richness_file)
  }
  
  city_results <- city_results %>%
    select(City, Threshold, Term, Estimate, Std_Error, t_value, p_value)
  
  city_richness_results <- city_richness_results %>%
    select(City, Threshold, Site, X, Estimator, Est_s.e., X95..Lower, X95..Upper, Total.Observations, Impervious)
  
  city_path <- paste0(richness_path, city_name, "/")
  
  write.csv(city_results, paste0(city_path, "lm_results_all_thresholds.csv"), quote=F, row.names=F)
  
  write.csv(city_richness_results, paste0(city_path, "richness_all_thresholds.csv"), quote=F, row.names=F)
}

```

# Collect all results across cities

```{r}

all_results <- data.frame(
  City = character(),
  Threshold = numeric(),
  Term = character(),
  Estimate = numeric(),
  Std_Error = numeric(),
  t_value = numeric(),
  p_value = numeric()
)

all_richness_results <- data.frame(
  City = character(),
    Threshold = numeric(),
    Site = character(),
    X = numeric(),
    Estimator = numeric(),
    Est_s.e. = numeric(),
    X95..Lower = numeric(),
    X95..Upper = numeric(),
    Total.Observations = numeric(),
    Impervious = numeric()
)

for (city_name in cities) {
  city_path <- paste0(richness_path, city_name, "/")
  
  results_path <- paste0(city_path, "lm_results_all_thresholds.csv")
  
  richness_file_path <- paste0(city_path, "richness_all_thresholds.csv")
  
  if (file.exists(results_path) == F) {
    next
  }
  
  if (file.exists(richness_file_path) == F) {
    next
  }
  
  cur_file <- read.csv(results_path)
  
  all_results <- rbind(all_results, cur_file)
  
  cur_richness_file <- read.csv(richness_file_path)
  
  all_richness_results <- rbind(all_richness_results, cur_richness_file)
}

all_results <- all_results %>%
  select(City, Threshold, Term, Estimate, Std_Error, t_value, p_value)

all_richness_results <- all_richness_results %>%
  select(City, Threshold, Site, X, Estimator, Est_s.e., X95..Lower, X95..Upper, Total.Observations, Impervious)

write.csv(all_results, paste0(richness_path, "lm_results_all_cities.csv"), quote=F, row.names=F)

write.csv(all_richness_results, paste0(richness_path, "richness_all_cities.csv"), quote=F, row.names=F)

```
